---
title: "transcriptomics_day2_kimia"
author: "Kimia"
format: html
editor: visual
toc: true
---

# **The basics of transcriptomics pt2**

Counting how many reads are mapped, the more reads mapped the more they are.

## **Part 1: Environmental Check & Collaboration**

### **1. Activation & R Libraries**

```{bash}
#| eval: false
conda activate
```

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(systemPipeRdata)
library(DT) # For interactive tables
library(Rsubread)
library(stringr)
```

### **2. Git Collaboration**

## **Part 2: The Metadata Architecture**

### **1. Loading the Data**

```{r}
# Load the master metadata table provided by systemPipeRdata
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, sep = "\t")
```

### **2. Inspecting the Structure**

```{r}
glimpse(meta_data)
datatable(meta_data)
```

### **3. The Wrangling Challenge**

```{r}
# Create a clean 'Type' column based on patterns in SampleLong
# case_when allows us to create multiple conditions in a single step
# grepl returns TRUE if the pattern (e.g., "Mock") is found in the string
meta_data <- meta_data %>% 
  mutate(Type = case_when(
    grepl("Mock", SampleLong) ~ "Mock",
    grepl("Avr", SampleLong) ~ "Avirulent",
    grepl("Vir", SampleLong) ~ "Virulent"
  ))
```

```{r}
meta_data |>
  select(SampleName, SampleLong, Type) |>
  head()
```

### **4. The Filter Challenge**

-   **Partner 1:** You are responsible for **Mock** and **Virulent**.

```{r}
my_targets <- meta_data %>%
  filter(Type %in% c("Mock", "Virulent"))
```

```{r}
print(my_targets$SampleName)
```

## **Part 3: The Pipeline Architecture**

### **1. Driving the Terminal with `system2()`**

The `system2()` function is R’s way of sending a command to your computer’s terminal.

**Anatomy of `system2("command", args = "...")`:**

1.  **The Command:** The name of the software (e.g., `"ls"`, `"hisat2"`, `"samtools"`).

2.  **Args:** A character vector of flags and files you would usually type in the terminal. If you need to provide multiple arguments, you must combine them into a single character vector using the `c()` function (e.g., `args = c("-l", "-h")`).

**Try it yourself:** Run these examples in your console to see how R communicates with your OS.

```{r}
# Example 1: List files in your current directory with multiple flags
system2("ls", args = c("-F", "-a"))

# Example 2: Check the current date/time from the system clock
system2("date")
```

```{r}
# Practical Question: Write a system2 call to check the version of samtools (--version)
# Write this in your .qmd
system2("samtools", args = "--version")
```

### **2. The Anatomy of a `for` loop**

```{r}
# The Collection: A sequence of numbers from 1 to the number of rows
for (i in 1:nrow(my_targets)) {
  # 1. Print the current value of the iterator i (the row number)
  print(paste("Currently processing row number:", i))
  
  # 2. Look up the SampleName for row 'i' and print it
  print(my_targets$SampleName[i])
}
```

```{r}

for (i in 1:nrow(my_targets)) {
  print(paste("Currently processing row number:", i))

    s_name <- my_targets$SampleName[i]
    print(s_name)
}
```

### **3. Building Paths with `paste0()` and `basename()`**

```{r}
for (i in 1:nrow(my_targets)) {
 print(paste0("data/data_kimia/trimmed/", str_remove(basename(my_targets$FileName1[i]), "gz"), "trimmed.fastq"))
}
```

### **4. Putting it Together: The Alignment Loop**

We will now build the “Engine” of our pipeline. The logic follows these steps:

1.  **Assign** the sample name to a variable.

2.  **Define** the input and output file paths.

3.  **Run** HISAT2 (Alignment).

4.  **Run** Samtools Sort (Compression).

5.  **Run** Samtools Index (Fast Retrieval).

6.  **Cleanup** (Delete the large SAM files).

```{r}
index_prefix <- "data/data_kimia/reference/athal_index"

for (i in 1:nrow(my_targets)) {
  
  # Step 1: Assign the SampleName to the variable s_name
  s_name <- my_targets$SampleName[i]
  
  # Step 2: Define path to input file and assign it to input_fq
  # YOUR TASK: Use the basename() and paste0() logic you practiced in Step 3!
 input_fq <- paste0("data/data_kimia/trimmed/", str_remove(basename(my_targets$FileName1[i]), "gz"), "trimmed.fastq")
 
  # Step 3: Given Paths to sam and bam output files
  sam_out  <- paste0("data/data_kimia/alignments/samfiles/", s_name, ".sam")
  bam_out  <- paste0("data/data_kimia/alignments/bamfiles/", s_name, ".sorted.bam")
  
  # Step 4: RUN HISAT2 (Alignment)
  # Pay attention to args: -p 8, -x [index], -U [input], -S [output]
  print(paste("Mapping:", s_name))
 system2("hisat2", args = c("-x", index_prefix, "-U", input_fq, "-p 8", "-S", sam_out))
  # system2("hisat2", args = c(...))
  
  # Step 5: RUN Samtools sort (Compression and ordering)
  # Pay attention to args: sort, -@ 8, -o [output_bam], [input_sam]
  print(paste("Sorting:", s_name))
  system2("samtools", args = c ("sort", "-@ 8", "-o", bam_out, sam_out))
  # system2("samtools", args = c(...))
  
  # Step 6: Index the bam file (Fast retrieval)
  # Pay attention to args: index, [input_bam]
  # system2("samtools", args = c(...))
  system2("samtools", args = c ("index", bam_out))
  
  # Step 7: DELETE THE SAM (Space management!)
  file.remove(sam_out)
}
```

### **5. Verification & Summary**

All good! :)))

## **Part 4: Generating the Count Matrix**

### **1. What is Rsubread?**

To perform this task, we use the `Rsubread` package. This is an R-wrapper for the **Subread** suite, a high-performance software package designed specifically for mapping and counting genomic reads.

### **2. Task: Running the Counts**

```{r}
# 1. Gather your BAM files
my_bams <- list.files("data/data_kimia/alignments/bamfiles", pattern = "\\.sorted\\.bam$", full.names = TRUE)

# 2. Reference the GTF structural map
gtf_file <- "data/data_kimia/reference/GCF_000001735.4_TAIR10.1_genomic.gtf"

# 3. Generate the counts
fc_results <- featureCounts(files = my_bams,
                            annot.ext = gtf_file,
                            isGTFAnnotationFile = TRUE,
                            GTF.featureType = "exon",
                            GTF.attrType = "gene_id",
                            nthreads = 8)

# 4. Extract the raw count matrix
raw_counts <- fc_results$counts
```

```{r}
datatable(raw_counts)
```

## **Part 5: Data Harmonization**

### **Task 3: The Tidyverse Cleanup**

```{r}
# Clean the names by removing file paths and extensions using stringr
colnames(raw_counts) <- colnames(raw_counts) %>%
  str_remove("data.alignments.") %>%
  str_remove(".sorted.bam")
```

#### **The Final Sanity Check**

```{r}
# THE SANITY CHECK: This MUST return TRUE.
all(colnames(raw_counts) %in% my_targets$SampleName)
```

# **Day 2 Wrap-Up: From Reads to Results**

Let’s recap what you achieved today and where it fits in the big picture:

1.  **Metadata Wrangling:** You took raw experimental notes and transformed them into a tidy, computer-readable format using `mutate` and `case_when`. This is the foundation of all downstream group comparisons.

2.  **Pipeline Automation:** You graduated from running single terminal commands to building a scalable **Bioinformatics Pipeline**. By mastering `system2` and `for` loops, you can now process 10, 100, or 1000 samples with the same effort.

3.  **Genomic Mapping:** You aligned short sequences to a reference genome, paying special attention to splice-aware junctions—a critical step in eukaryotes.

4.  **Quantification:** Using `featureCounts`, you turned abstract sequence alignments into a **Gene Expression Matrix**. This is the digital equivalent of counting which “instructions” (mRNA) were active in the cell.

# **Final Submission**

```{r}
write.csv(raw_counts, "data/MockVsVirulent_Specific_Counts.csv", row.names = TRUE)
```
