---
title: "transcriptomics_day4_kimia"
author: "Kimia"
format: html
editor: visual
toc: true
---

# **The basics of transcriptomics pt4**

## **Part 0: Computational Environment**

```{r}
#| message: false
#| warning: false
# Check and install missing critical packages if they don't exist
if (!requireNamespace("org.At.tair.db", quietly = TRUE)) BiocManager::install("org.At.tair.db", update = FALSE)
if (!requireNamespace("viridis", quietly = TRUE)) install.packages("viridis")

# Define the required packages
# clusterProfiler: For GO enrichment
# org.At.tair.db: The Arabidopsis annotation database
# viridis: For colorblind-friendly palettes
pkgs <- c("tidyverse", "DESeq2", "pheatmap", "EnhancedVolcano", 
          "clusterProfiler", "org.At.tair.db", "RColorBrewer", "viridis")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)
```

## **Part 1: The Volcano Plot**

1.  Generate a Volcano plot for your 1h vs. 6h results.

2.  Visually identify if the “mountain” of genes looks broader or narrower than the 12h comparison.

**Bonus:** Do it for 6h vs 12h

### 12h vs. 1h

```{r}
EnhancedVolcano(res_1h_12h,
                lab = rownames(res_1h_12h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 12h vs. 1h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

### 1h vs. 6h

```{r}
EnhancedVolcano(res_1h_6h,
                lab = rownames(res_1h_6h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 1h vs. 6h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

### 6h vs. 12h

```{r}
EnhancedVolcano(res_6h_12h,
                lab = rownames(res_6h_12h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 6h vs. 12h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

## **Part 2: The Heatmap and Z-Scores**

```{r}
vsd2 <- varianceStabilizingTransformation(dds2, blind =FALSE)
vsd_mat2 <- assay(vsd2) # Extract the matrix of transformed values
```

### **3. Task: Visualizing the Top Responders (with Accessible Design)**

```{r}
sig_1h_12h <- read.csv("data/data_kimia/sig_1h_12h_genes.csv", row.names = 1)
sig_1h_6h <- read.csv("data/data_kimia/sig_1h_6h_genes.csv", row.names = 1)
sig_6h_12h <- read.csv("data/data_kimia/sig_6h_12h_genes.csv", row.names = 1)
```

#### 1h vs. 12h

```{r}
# 1. Extract the VST-transformed counts for your significant 1h vs 12h genes
# Assuming your filtered object is named 'sig_1h_12h'
plot_mat_12h <- assay(vsd2)[c(rownames(sig_1h_12h),row.names(sig_1h_6h)[-which(row.names(sig_1h_6h)%in%row.names(sig_1h_12h))]), ]

# 2. Create the Heatmap
pheatmap(plot_mat_12h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```

#### 1h vs. 6h

```{r}
plot_mat_1h<- assay(vsd2)[rownames(sig_1h_6h), ]

# 2. Create the Heatmap
pheatmap(plot_mat_1h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 1h vs 6h")
```

#### 12h vs. 6h

```{r}
plot_mat_12h_6h <- assay(vsd2)[rownames(sig_6h_12h), ]

# 2. Create the Heatmap
pheatmap(plot_mat_12h_6h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 6h")
```

### **Bonus: The “Human Readable” Heatmap**

#### 1h vs. 12h

Found LHY.

```{r}
# 1. Map TAIR IDs to Symbols using the annotation database
gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_12h), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_12h) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_12h))

# 3. Now re-run your pheatmap() call with plot_mat_12h!
pheatmap(plot_mat_12h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
rownames(plot_mat_12h)
```

#### 1h vs. 6h 

```{r}
gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_1h), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_1h) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_1h))

# 3. Now re-run your pheatmap() call with plot_mat_12h!
pheatmap(plot_mat_1h)
rownames(plot_mat_1h)
```

#### 6h vs. 12h

```{r}
gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_12h_6h), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_12h_6h) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_12h_6h))

# 3. Now re-run your pheatmap() call with plot_mat_12h!
pheatmap(plot_mat_12h_6h)
rownames(plot_mat_12h_6h)
```

## **Part 3: Gene Ontology (GO) Enrichment**

### **2. Running enrichGO**

#### 1h vs. 12h

```{r}
sig_1h_12h_up <- sig_1h_12h %>%
  as.data.frame() %>%
  filter(padj < 0.2 & log2FoldChange >= 1) %>%
  rownames()

ego_1h_12h <- enrichGO(gene       = row.names (sig_1h_12h),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_1h_12h, showCategory=20) + ggtitle("GO Enrichment: 1h vs. 12h")
ggplot2::ggsave(
  filename = "data/data_kimia/dotplot_1h_12h.png",
  plot = enrichplot::dotplot(ego_1h_12h, showCategory=20) + 
    ggplot2::ggtitle("GO Enrichment: 1h vs. 12h"),
  width = 8, 
  height = 6
)

```

#### 1h vs. 6h 

```{r}
ego_1h_6h <- enrichGO(gene       = row.names(sig_1h_6h),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_1h_6h, showCategory=20) + ggtitle("GO Enrichment: 1h vs. 6h")
```

#### 6h vs. 12h 

```{r}
sig_6h_12h_up <- sig_6h_12h %>%
  as.data.frame() %>%
  filter(padj < 0.2 & log2FoldChange >= 1) %>%
  rownames()

sig_6h_12h_up <- sig_6h_12h %>%
  as.data.frame() %>%
  filter(padj < 0.2 & log2FoldChange >= 1) %>%
  rownames()

ego_6h_12h <- enrichGO(gene       = row.names(sig_6h_12h),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

ego_6h_12h <- enrichGO(gene       = sig_6h_12h_up,
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_6h_12h, showCategory=20) + ggtitle("GO Enrichment: 6h vs. 12h")
```

### **Pro Tip:** Custom Plotting with geom_point

#### 1h vs. 12h 

```{r}
# 1. Convert to a dataframe
ego_1h_12h_df <- as.data.frame(ego_1h_12h) %>%
  head(20) %>%
  # Secret trick: GeneRatio is a string "5/40". 
  # We use parse/eval to turn it into a decimal number (0.125).
  mutate(GeneRatioNumeric = sapply(GeneRatio, function(x) eval(parse(text=x))))

# 2. Plot with geom_point
ggplot(ego_1h_12h_df, aes(x = GeneRatioNumeric, y = reorder(Description, GeneRatioNumeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(x = "Gene Ratio", y = "Biological Process", 
       title = "Custom GO Enrichment Plot",
       size = "Gene Count", color = "Adj. P-value")
```

#### 1h vs. 6h 

```{r}
# 1. Convert to a dataframe
ego_1h_6h_df <- as.data.frame(ego_1h_6h) %>%
  head(20) %>%
  # Secret trick: GeneRatio is a string "5/40". 
  # We use parse/eval to turn it into a decimal number (0.125).
  mutate(GeneRatioNumeric = sapply(GeneRatio, function(x) eval(parse(text=x))))

# 2. Plot with geom_point
ggplot(ego_1h_6h_df, aes(x = GeneRatioNumeric, y = reorder(Description, GeneRatioNumeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(x = "Gene Ratio", y = "Biological Process", 
       title = "Custom GO Enrichment Plot",
       size = "Gene Count", color = "Adj. P-value")
```

#### 6h vs. 12h 

```{r}
ego_6h_12h_df <- as.data.frame(ego_6h_12h) %>%
  head(20) %>%
  # Secret trick: GeneRatio is a string "5/40". 
  # We use parse/eval to turn it into a decimal number (0.125).
  mutate(GeneRatioNumeric = sapply(GeneRatio, function(x) eval(parse(text=x))))

# 2. Plot with geom_point
ggplot(ego_6h_12h_df, aes(x = GeneRatioNumeric, y = reorder(Description, GeneRatioNumeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(x = "Gene Ratio", y = "Biological Process", 
       title = "Custom GO Enrichment Plot",
       size = "Gene Count", color = "Adj. P-value")
```

## **Part 4: The Final Synthesis**

## **Day 4 Wrap-Up: From Data to Knowledge**

# **Final Submission**
