---
title: "transcriptomics_day2_ankur"
format: html
---

```{r}
library(tidyverse)
library(systemPipeRdata)
library(DT) # For interactive tables
library(Rsubread)
library(stringr)
```

### Loading the Data

```{r}
# Load the master metadata table provided by systemPipeRdata
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, sep = "\t")
```

### Inspecting the Structure

```{r}
glip <- glimpse(meta_data)
hea <- head(meta_data)
```
```{r}
datatable(meta_data) 
```
```{r}
# Create a clean 'Type' column based on patterns in SampleLong
# case_when allows us to create multiple conditions in a single step
# grepl returns TRUE if the pattern (e.g., "Mock") is found in the string
meta_data <- meta_data %>% 
  mutate(Type = case_when(
    grepl("Mock", SampleLong) ~ "Mock",
    grepl("Avr", SampleLong) ~ "Avirulent",
    grepl("Vir", SampleLong) ~ "Virulent"
  ))
```

```{r}
meta_data |>
select(SampleName, SampleLong, Type) |> 
  head()
```

```{r}
# Filtering
my_targets <- meta_data |>
  filter(Type %in% c("Mock", "Avirulent"))

```

```{r}
#Sanity Check
my_targets |>
  select(SampleName) |>
  head()
```

### Part 3: The Pipeline Architecture

### Pre-requsite 


```{r}
# Example 1: List files in your current directory with multiple flags
system2("ls", args = c("-F", "-a"))

# Example 2: Check the current date/time from the system clock
system2("date")

# Practical Question: Write a system2 call to check the version of samtools (--version)
# Write this in your .qmd

system2("samtools", args = "--version")
```


```{r}
# The Collection: A basket of fruits
fruits <- c("Apple", "Banana", "Cherry")

# The Loop: 
for (f in fruits) {
  print(paste("I like", f))
}
```

```{r}
# The Collection: A sequence of numbers from 1 to the number of rows
for (i in 1:nrow(my_targets)) {
  # 1. Print the current value of the iterator i (the row number)
  print(paste("Currently processing row number:", i))
  
  # 2. Look up the SampleName for row 'i' and print it
  print(my_targets$SampleName[i])
  
}
```

```{r}
# The Collection: A sequence of numbers from 1 to the number of rows
for (i in 1:nrow(my_targets)) {
  # 1. Print the current value of the iterator i (the row number)
  print(paste("Currently processing row number:", i))
  
  # 2. Look up the SampleName for row 'i' and print it
  s_name <- print(my_targets$SampleName[i])
  
}
```


```{r}
for (i in 1:nrow(my_targets)) {
  print(paste0("/data/data_ankur/trimmed/", str_remove(basename(my_targets$FileName1[i]),"gz"), "trimmed.fastq"))
  
}
```

### Pipeline

```{r}
dir.create("data/data_ankur/alignments", showWarnings = FALSE)
index_prefix <- "data/data_ankur/reference/athal_index"

for (i in 1:nrow(my_targets)) {
  
  # Step 1: Assign the SampleName to the variable s_name
  s_name <- my_targets$SampleName[i]
  
  # Step 2: Define path to input file and assign it to input_fq
  # YOUR TASK: Use the basename() and paste0() logic you practiced in Step 3!
  
  input_fq <- paste0("data/data_ankur/trimmed/", str_remove(basename(my_targets$FileName1[i]),"gz"), "trimmed.fastq")
  

  
  # Step 3: Given Paths to sam and bam output files
  sam_out  <- paste0("data/data_ankur/alignments/", s_name, ".sam")
  bam_out  <- paste0("data/data_ankur/alignments/", s_name, ".sorted.bam")
  
  # Step 4: RUN HISAT2 (Alignment)
  # Pay attention to args: -p 8, -x [index], -U [input], -S [output]
  print(paste("Mapping:", s_name))
  # system2("hisat2", args = c(...))
  system2("hisat2", args = c("-p 8","-x", index_prefix,"-U", input_fq, "-S", sam_out))
  
  # Step 5: RUN Samtools sort (Compression and ordering)
  # Pay attention to args: sort, -@ 8, -o [output_bam], [input_sam]
  print(paste("Sorting:", s_name))
  # system2("samtools", args = c(...))
  system2("samtools", args = c("sort", "-o", bam_out, sam_out))
  
  # Step 6: Index the bam file (Fast retrieval)
  # Pay attention to args: index, [input_bam]
  # system2("samtools", args = c(...))
  system2("samtools", args = c("index", bam_out))
  
  # Step 7: DELETE THE SAM (Space management!)
  file.remove(sam_out)
}

```
```{bash}
du -h data/data_ankur/alignments/*.bam
```
### Running the Counts

```{r}
# 1. Gather your BAM files
my_bams <- list.files("data/data_ankur/alignments", pattern = "\\.sorted\\.bam$", full.names = TRUE)

# 2. Reference the GTF structural map
gtf_file <- "data/data_ankur/reference/GCF_000001735.4_TAIR10.1_genomic.gtf"

# 3. Generate the counts
fc_results <- featureCounts(files = my_bams,
                            annot.ext = gtf_file,
                            isGTFAnnotationFile = TRUE,
                            GTF.featureType = "exon",
                            GTF.attrType = "gene_id",
                            nthreads = 8)

# 4. Extract the raw count matrix
raw_counts <- fc_results$counts
```

## Data Harmonization

```{r}
# Clean the names by removing file paths and extensions using stringr
colnames(raw_counts) <- colnames(raw_counts) %>%
  str_remove("data.alignments.") %>%
  str_remove(".sorted.bam")
```


```{r}
# THE SANITY CHECK: This MUST return TRUE.
all(colnames(raw_counts) %in% my_targets$SampleName)
```

```{r}
write.csv(raw_counts, "data/mock_vs_avr.csv", row.names = TRUE)
```


## Reflection: 

1. we delete sam in the end of loop to save disk space 

2. While the aligner (e.g., STAR, HISAT2) determines where a read maps on the genome, the GTF (Gene Transfer Format) file provides the annotated, hierarchical structure of the transcriptome that the aligner does not inherently know.
