---
title: "transcriptomics_day1_p1.qmd"
author: "Kimia"
format: html
editor: visual
toc: true
---

# **The basics of transcriptomics pt1**

## **Part 1: The Digital Mise-en-Place**

### **The Environment Audit**

```{r}
#| message: false
#| warning: false

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

cran_pkgs <- c("tidyverse", "pheatmap")
bioc_pkgs <- c("DESeq2", "clusterProfiler", "Rsubread", "systemPipeRdata", "Rqc", "QuasR")

new_pkgs <- c(cran_pkgs, bioc_pkgs)[!(c(cran_pkgs, bioc_pkgs) %in% installed.packages()[,"Package"])]
if(length(new_pkgs)) BiocManager::install(new_pkgs, update = FALSE)

lapply(c(cran_pkgs, bioc_pkgs), library, character.only = TRUE)
```

```{bash}
#| eval: false
conda install -c bioconda -c samtools hisat2
```

## **Part 2: Data Acquisition**

### **1. The Reference & Reads**

```{bash}
#| eval: false
mkdir -p data/data_kimia/reference data/raw data/trimmed reports

# Download TAIR10.1
wget -nc https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/735/GCF_000001735.4_TAIR10.1/GCF_000001735.4_TAIR10.1_genomic.fna.gz -P data/reference/
wget -nc https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/735/GCF_000001735.4_TAIR10.1/GCF_000001735.4_TAIR10.1_genomic.gtf.gz -P data/reference/
gunzip data/reference/*.gz
```

```{r}
library(systemPipeRdata)
fq_path <- systemPipeRdata::pathList()$fastqdir
file.copy(list.files(fq_path, full.names = TRUE), "data/raw/")
```

```{bash}
#| eval: false
gunzip data/raw/*.gz
```

## **Part 3: The Theory - Beyond the Blueprint**

## **Part 4: The Practice - R-Based QC and Trimming**

### **Task 1: Generate the QC Report**

```{r}
#| message: false
#| warning: false

library(Rqc)
raw_files <- list.files("data/raw", pattern = ".fastq", full.names = TRUE)
qc_res <- rqc(path = "data/raw", pattern = ".fastq", outdir = "reports/rqc_raw")
```

### **Task 2: Strict Filtering with QuasR**

```{r}
#| message: false
#| warning: false

library(QuasR)

# Filter and output to 'trimmed' folder with new names
preprocessReads(filename = raw_files,
                outputFilename = paste0("data/data_kimia/trimmed_kimia/", basename(raw_files), ".trimmed.fastq"),
                truncateEndBases = 3, 
                minLength = 30, 
                maxN = 1)
```

## **Part 5: Mapping Strategy - Splicing Matters**

## **Part 6: Building the Index via `system2()`**

```{r}
#| message: false
#| warning: false

# Reference and Index paths
ref_genome <- "data/reference/GCF_000001735.4_TAIR10.1_genomic.fna"
index_prefix <- "data/reference/athal_index"

# Drive the terminal using R: 8 CPUs for speed
system2("hisat2-build", args = c("-p 8", ref_genome, index_prefix))
```

## **Bonus Challenges**

### **Bonus 1: Planning the Map**

```{bash}
#| eval: false
hisat2 -h

```

### **Bonus 2: Advanced `system2` Design**

### Reflection

Explain in your own words the difference between a `.bai` index and a HISAT2 index. Why can’t we use a `.bai` to map reads?
