---
title: "transcriptomics_day4_kimia"
author: "Kimia"
format: html
editor: visual
toc: true
---

# **The basics of transcriptomics pt4**

## **Part 0: Computational Environment**

```{r}
#| message: false
#| warning: false
# Check and install missing critical packages if they don't exist
if (!requireNamespace("org.At.tair.db", quietly = TRUE)) BiocManager::install("org.At.tair.db", update = FALSE)
if (!requireNamespace("viridis", quietly = TRUE)) install.packages("viridis")

# Define the required packages
# clusterProfiler: For GO enrichment
# org.At.tair.db: The Arabidopsis annotation database
# viridis: For colorblind-friendly palettes
pkgs <- c("tidyverse", "DESeq2", "pheatmap", "EnhancedVolcano", 
          "clusterProfiler", "org.At.tair.db", "RColorBrewer", "viridis")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)
```

## **Part 1: The Volcano Plot**

1.  Generate a Volcano plot for your 1h vs. 6h results.

2.  Visually identify if the “mountain” of genes looks broader or narrower than the 12h comparison.

**Bonus:** Do it for 6h vs 12h

### 12h vs. 1h

```{r}
EnhancedVolcano(res_1h_12h,
                lab = rownames(res_1h_12h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 12h vs. 1h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

### 1h vs. 6h

```{r}
EnhancedVolcano(res_1h_6h,
                lab = rownames(res_1h_6h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 1h vs. 6h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

### 6h vs. 12h

```{r}
EnhancedVolcano(res_6h_12h,
                lab = rownames(res_6h_12h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 6h vs. 12h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

## **Part 2: The Heatmap and Z-Scores**

```{r}
vsd2 <- varianceStabilizingTransformation(dds2, blind =FALSE)
vsd_mat2 <- assay(vsd2) # Extract the matrix of transformed values
```

### **3. Task: Visualizing the Top Responders (with Accessible Design)**

```{r}
sig_1h_12h <- read.csv("data/data_kimia/sig_1h_12h_genes.csv", row.names = 1)
sig_1h_6h <- read.csv("data/data_kimia/sig_1h_6h_genes.csv", row.names = 1)
sig_6h_12h <- read.csv("data/data_kimia/sig_6h_12h_genes.csv", row.names = 1)
```

#### 1h vs. 12h

```{r}
# 1. Extract the VST-transformed counts for your significant 1h vs 12h genes
# Assuming your filtered object is named 'sig_1h_12h'
plot_mat_12h <- assay(vsd2)[rownames(sig_1h_12h), ]

# 2. Create the Heatmap
pheatmap(plot_mat_12h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```

#### 1h vs. 6h

```{r}
plot_mat_1h<- assay(vsd2)[rownames(sig_1h_6h), ]

# 2. Create the Heatmap
pheatmap(plot_mat_1h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 1h vs 6h")
```

#### 12h vs. 6h

```{r}
plot_mat_12h_6h <- assay(vsd2)[rownames(sig_6h_12h), ]

# 2. Create the Heatmap
pheatmap(plot_mat_12h_6h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 6h")
```
