---
title: "transcriptomics_day2_Krish.qmd"
author: "Krish Desai"
format: html
editor: visual
---

# **Part 1: Environmental Check & Collaboration**

### **1. Activation & R Libraries**

```{bash}
conda activate
```

Load the necessary R libraries for today if you restarted your R session from Day 1.

```{r}
#| echo: false
library(tidyverse)
library(systemPipeRdata)
library(DT) # For interactive tables
library(Rsubread)
library(stringr)
```

# **Part 2: The Metadata Architecture**

### **1. Loading the Data**

```{r}
# Load the master metadata table provided by systemPipeRdata
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, sep = "\t")
```

### **2. Inspecting the Structure**

```{r}
head(meta_data)
glimpse(meta_data)

datatable(meta_data)
```

### **3. The Wrangling Challenge**

```{r}
# Create a clean 'Type' column based on patterns in SampleLong
# case_when allows us to create multiple conditions in a single step
# grepl returns TRUE if the pattern (e.g., "Mock") is found in the string
meta_data <- meta_data %>% 
  mutate(Type = case_when(
    grepl("Mock", SampleLong) ~ "Mock",
    grepl("Avr", SampleLong) ~ "Avirulent",
    grepl("Vir", SampleLong) ~ "Virulent"
  ))

meta_data|>
  select(SampleName, SampleLong, Type )|>
  head()

```

### **4. The Filter Challenge**

```{r}
my_targets <- meta_data %>%
  filter(Type %in% c("Virulent", "Avirulent"))

print(my_targets$SampleName)
```

# **Part 3: The Pipeline Architecture**

### **1. Driving the Terminal with `system2()`**

```{r}
# Example 1: List files in your current directory with multiple flags
system2("ls", args = c("-F", "-a"))

# Example 2: Check the current date/time from the system clock
system2("date")

# Practical Question: Write a system2 call to check the version of samtools (--version)
# Write this in your .qmd

system2("samtools", args = c("--version"))
```

### **2. The Anatomy of a `for` loop**

```{r}
# The Collection: A sequence of numbers from 1 to the number of rows
for (i in 1:nrow(my_targets)) {
  # 1. Print the current value of the iterator i (the row number)
  print(paste("Currently processing row number:", i))
  
  # 2. Look up the SampleName for row 'i' and print it
  print(my_targets$SampleName[i])
  
  s_name <- my_targets$SampleName
  
  print(s_name)
}
```

### **3. Building Paths with `paste0()` and `basename()`**

```{r}
for (i in 1:nrow(my_targets)) {
  print(paste0("data/Data_Krish/trimmed/", str_remove(basename(my_targets$FileName1[i]), "gz"), "trimmed.fastq"))
}
```

### **4. Putting it Together: The Alignment Loop**

```{r}
index_prefix <- "data/Data_Krish/reference/athal_index"

for (i in 1:nrow(my_targets)) {
  
  # Step 1: Assign the SampleName to the variable s_name
  s_name <- my_targets$SampleName[i]
  
  # Step 2: Define path to input file and assign it to input_fq
  # YOUR TASK: Use the basename() and paste0() logic you practiced in Step 3!
    input_fq <- paste0("data/Data_Krish/trimmed/", str_remove(basename(my_targets$FileName1[i]), "gz"), "trimmed.fastq")
    
  # Step 3: Given Paths to sam and bam output files
  sam_out  <- paste0("data/Data_Krish/alignment/", s_name, ".sam")
  bam_out  <- paste0("data/Data_Krish/alignment/", s_name, ".sorted.bam")
  
  # Step 4: RUN HISAT2 (Alignment)
  # Pay attention to args: -p 8, -x [index], -U [input], -S [output]
  print(paste("Mapping:", s_name))
  # system2("hisat2", args = c(...))
  system2("hisat2" , args = c("-x" , index_prefix,  "-U", input_fq, "-p 8", "-S", sam_out))
  
  # Step 5: RUN Samtools sort (Compression and ordering)
  # Pay attention to args: sort, -@ 8, -o [output_bam], [input_sam]
  print(paste("Sorting:", s_name))
  # system2("samtools", args = c(...))
  system2("samtools", args = c("sort", "-@ 8", "-o", bam_out, sam_out ))
  
  
  
  
  # Step 6: Index the bam file (Fast retrieval)
  # Pay attention to args: index, [input_bam]
  # system2("samtools", args = c(...))
  system2("samtools", args = c("index", bam_out))
  
  # Step 7: DELETE THE SAM (Space management!)
  file.remove(sam_out)
}
```

### **5. Verification & Summary**

```{bash}
du -h data/Data_Krish/alignment/*.bam
```

# **Part 4: Generating the Count Matrix**
