---
title: "ranscriptomics_day3_Krish"
author: "Krish Desai"
format: html
editor: visual
---

# **Part 0: Computational Environment**

```{bash}
conda create transcriptomics_env python=3.8
```

```{r}
#| echo: false


# Define the required packages
pkgs <- c("tidyverse", "DESeq2", "DT", "pheatmap", "RColorBrewer", "EnhancedVolcano", "ape")
BiocManager::install('EnhancedVolcano')

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)

```

# **Part 2: The Experimental Design Matrix**

### **1. Creating the colData**

```{r}
counts_master <- read.csv("data/mock_vs_avr_vs_vir.csv", row.names = 1)
# Create a metadata frame
# Ensure the levels are set so 'Mock' is the reference (the baseline)
colData <- data.frame(
  row.names = colnames(counts_master),
  Condition = factor(rep(c("Virulent", "Avirulent", "Mock"), each = 6), 
                     levels = c("Virulent", "Avirulent", "Mock")),
  Time = factor(rep(rep(c("12h", "1h", "6h"), each = 2), 3))
)

```

### **2. Task: Metadata Expansion**

```{r}
colData <- colData %>%
  mutate(type = factor(rep("single-read")))

all(rownames(colData) == colnames(counts_master))
```

# **Part 3: Introduction to DESeq2**

### **2. Constructing the DESeq Object**

```{r}
# Create the dds object
dds <- DESeqDataSetFromMatrix(countData = round(counts_master),
                              colData = colData,
                              design = ~ Condition)

dds2 <- DESeqDataSetFromMatrix(countData = round(counts_master),
                              colData = colData,
                              design = ~ Time)
```

# **Part 4: Exploratory Data Analysis: Hierarchical Clustering**

### **3. Task: Building the Dendrogram**

```{r}
# 1. Transform the data to stabilize variance
vsd <- varianceStabilizingTransformation(dds)
vsd2 <- varianceStabilizingTransformation(dds2)
vsd_mat <- assay(vsd) # Extract the matrix of transformed values
vsd_mat2 <- assay(vsd2) # Extract the matrix of transformed values
# 2. Calculate the distance between samples
# We transpose the matrix with t() because dist() calculates distance between rows
sample_dist <- dist(t(vsd_mat))
sample_dist2 <- dist(t(vsd_mat2))

# 3. Perform Hierarchical Clustering
clusters <- hclust(sample_dist)
clusters2 <- hclust(sample_dist2)


# 4. Plot the results using the ape package for a cleaner phylogenetic look
# We convert the hclust object to a phylo object for professional plotting.
# We map 'Condition' to 'tip.label' so the tree is easy to interpret.
plot.phylo(as.phylo(clusters), 
           tip.label = colData$Condition, 
           type = "p", 
           edge.col = "red", 
           edge.width = 2,
           no.margin = TRUE)

plot.phylo(as.phylo(clusters2), 
           tip.label = colData$Time, 
           type = "p", 
           edge.col = "red", 
           edge.width = 2,
           no.margin = TRUE)
```

**Now You:** Run the clustering code in your `.qmd`.

1.  Look at the dendrogram. Do the replicates for the same condition (e.g., the two `Mock` samples at 1h) cluster together on the same small branches?

    No, they don't cluster together.

2.  Do you see a clear split between the `Mock` group and the infected groups?

    No, they were randomly clustered

3.  If a sample is “out of place” (e.g., a Virulent sample sitting in the middle of the Mock branch), what might that suggest about the experiment

# **Part 5: The Main Event - Differential Expression**

### **1. Running the Pipeline**

```{r}
# Assign the output to a new object to keep your original dds safe
dds_results <- DESeq(dds)

dds_results2 <- DESeq(dds2)
```

### **2. Extracting Specific Results**

```{r}
# Extract results for Avirulent vs. Mock
res_avr_mock <- results(dds_results, contrast = c("Condition", "Avirulent", "Mock"), alpha = 0.2)
res_vir_mock <- results(dds_results, contrast = c("Condition", "Virulent", "Mock"), alpha = 0.2)
res_vir_avr <- results(dds_results, contrast = c("Condition", "Virulent", "Avirulent"), alpha = 0.2)




# Print the first few rows to see the table
print(res_avr_mock)
summary(res_avr_mock)
summary(res_vir_mock)
summary(res_vir_avr)

```

#### **Understanding the Results Table**

```{r}
# See a high-level summary of the results
res_1_12 <- results(dds_results2, contrast = c("Time", "1h", "12h"), alpha = 0.2)
res_1_6 <- results(dds_results2, contrast = c("Time", "1h", "6h"), alpha = 0.2)
res_6_12 <- results(dds_results2, contrast = c("Time", "6h", "12h"), alpha = 0.2)

print(res_1_12)

summary(res_1_12)
summary(res_1_6)
summary(res_6_12)
```

# **Part 6: Filtering and Ranking**

### **2. Automation with Custom Functions**

```{r}
# Function to filter and count DE genes
filter_and_count <- function(res_obj, comparison_name, fc_threshold = 2) {
  # 1. Remove rows with NAs in padj
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  
  # 2. Apply filters: |log2FC| >= log2(fc_threshold) and padj < 0.2
  # Note: 0.2 matches the alpha we set in our results() calls
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold) & res_filtered$padj < 0.2, ]
  
  # 3. Count up and down regulated
  up_regulated <- sum(sig_genes$log2FoldChange > 0)
  down_regulated <- sum(sig_genes$log2FoldChange < 0)
  
  # Return as a clean data frame
  return(data.frame(
    Comparison = comparison_name,
    Up_regulated = up_regulated,
    Down_regulated = down_regulated
  ))
}
```

```{r}
 results_summary <- rbind(
  filter_and_count(res_1_12, "1h vs 12h"),
  filter_and_count(res_1_6, "1h vs 6h"),
  filter_and_count(res_6_12, "6h vs 12h")
)

# Inspect the final summary table
print(results_summary)


 results_summary2<- rbind(
  filter_and_count(res_vir_mock, "Vir vs Mock"),
  filter_and_count(res_avr_mock, "Avr vs Mock"),
  filter_and_count(res_vir_avr, "Vir vs Avr")
)

# Inspect the final summary table
print(results_summary2)
```

```{r}
# Step 1: Pivot the data to long format
plot_data <- results_summary %>%
  pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count") %>%
  mutate(Regulation = factor(Regulation, levels = c("Up_regulated", "Down_regulated")))

# Step 2: Create a Bar Plot
# Hint: Use aes(x = Comparison, y = Count, fill = Regulation) 
# and geom_bar(stat = "identity", position = "dodge")

ggplot(plot_data, aes (x = Comparison, y = Count, fill = Regulation)) +
        geom_bar(stat = "identity", position = "dodge")
  
```

# **Day 3 Wrap-Up**

```{r}

# Function to filter and count DE genes
filter_and_count2 <- function(res_obj,fc_threshold = 2) {
  # 1. Remove rows with NAs in padj
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  
  # 2. Apply filters: |log2FC| >= log2(fc_threshold) and padj < 0.2
  # Note: 0.2 matches the alpha we set in our results() calls
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold) & res_filtered$padj < 0.2, ]
  return(sig_genes)

}
  


write.csv(filter_and_count2(res_1_12), "data/Data_Krish/sig_virulent_genes_filteredres_1_12.csv", row.names = TRUE)
write.csv(filter_and_count2(res_1_6), "data/Data_Krish/sig_virulent_genes2_filteredres_1_6.csv", row.names = TRUE)
write.csv(filter_and_count2(res_6_12), "data/Data_Krish/sig_virulent_genes_filteredres_6_12.csv", row.names = TRUE)
  
write.csv(res_1_12, "data/Data_Krish/sig_virulent_genes_res_1_12.csv", row.names = TRUE)
write.csv(res_1_6, "data/Data_Krish/sig_virulent_genes_res_1_6.csv", row.names = TRUE)
write.csv(res_6_12, "data/Data_Krish/sig_virulent_genes_res_6_12.csv", row.names = TRUE)


```
