---
title: "transcriptomics_day4_ankur"
format: html
---

## 1. Installation

```{r}
# Check and install missing critical packages if they don't exist
if (!requireNamespace("org.At.tair.db", quietly = TRUE)) BiocManager::install("org.At.tair.db", update = FALSE)
if (!requireNamespace("viridis", quietly = TRUE)) install.packages("viridis")

# Define the required packages
# clusterProfiler: For GO enrichment
# org.At.tair.db: The Arabidopsis annotation database
# viridis: For colorblind-friendly palettes
pkgs <- c("tidyverse", "DESeq2", "pheatmap", "EnhancedVolcano", 
          "clusterProfiler", "org.At.tair.db", "RColorBrewer", "viridis")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)
```

## 2. Loading Files

```{r}
# Load your dds_results and count matrix from Day 3


# Ensure these objects are in your environment before proceeding
print("Biological Discovery Environment: READY")
```


## 3. The Volcano Plot

```{r}
 EnhancedVolcano(res_1_12,
                lab = rownames(res_1_12),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 12h vs. 1h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```


```{r}
EnhancedVolcano(res_1_6,
                lab = rownames(res_1_6),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 1h vs. 6h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```


```{r}
EnhancedVolcano(res_6_12,
                lab = rownames(res_6_12),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 6h vs. 12h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

# The heatmap and Z-scores

```{r}
# 1. Extract the VST-transformed counts for your significant 1h vs 12h genes
# Assuming your filtered object is named 'sig_1h_12h'
plot_mat_12h <- assay(vsd)[rownames(b), ]

# 2. Create the Heatmap
pheatmap(plot_mat_12h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```
```{r}
# 1. Extract the VST-transformed counts for your significant 1h vs 12h genes
# Assuming your filtered object is named 'sig_11h_12h'

row.names(a)[-which(row.names(a)==row.names(b))]

plot_mat_112h <- assay(vsd)[c(rownames(b),row.names(a)[-which(row.names(a)%in%row.names(b))]), ]

# 2. Create the Heatmap
pheatmap(plot_mat_112h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```


```{r}
setdiff(rownames(b), rownames(a))
```

```{r}

plot_mat_6h <- assay(vsd)[rownames(a), ]

# 2. Create the Heatmap
pheatmap(plot_mat_6h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 6h vs 1h")
```

```{r}

plot_mat_612h <- assay(vsd)[rownames(c), ]

# 2. Create the Heatmap
pheatmap(plot_mat_612h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 6h")
```

```{r}
# 1. Map TAIR IDs to Symbols using the annotation database
gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_12h), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_12h) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_12h))

# 3. Now re-run your pheatmap() call with plot_mat_12h!
# 2. Create the Heatmap
pheatmap(plot_mat_12h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```
 
```{r}
# 1. Map TAIR IDs to Symbols using the annotation database
gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_112h), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_112h) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_112h))

print(rownames(plot_mat_112h))

# 2. Create the Heatmap
pheatmap(plot_mat_112h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")

```

```{r}
# 1. Extract the VST-transformed counts for your significant 1h vs 12h genes
# Assuming your filtered object is named 'sig_11h_12h'

row.names(a)[-which(row.names(a)==row.names(b))]

plot_mat_ah <- assay(vsd)[c(rownames(b),row.names(a)[-which(row.names(a)==row.names(b))]), ]

gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_ah), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_ah) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_ah))

print(rownames(plot_mat_ah))

# 2. Create the Heatmap
pheatmap(plot_mat_ah, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```
```{r}

sig_avr_up <- read.csv("data/fres_1_12_ankur.csv", row.names = 1)

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = base::rownames(sig_avr_up),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=20) + ggtitle("GO Enrichment: Avirulent Response")
```
```{r}
up_1_12_regulated <- sig_avr_up |>
  filter(log2FoldChange>=1)

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = base::rownames(up_1_12_regulated),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=20) + ggtitle("GO Enrichment: Avirulent Response")
```
```{r}
down_1_12_regulated <- sig_avr_up |>
  filter(log2FoldChange<=-1)

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = base::rownames(down_1_12_regulated),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=10) + ggtitle("GO Enrichment: Avirulent Response")
```
```{r}

sig_avr_up <- read.csv("data/fres_1_12_ankur.csv", row.names = 1)

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = base::rownames(sig_avr_up),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=20) + ggtitle("GO Enrichment: Avirulent Response")
```

```{r}
sig_avr_2 <- read.csv("data/fres_1_6_ankur.csv", row.names = 1)

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = base::rownames(sig_avr_up),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=20) + ggtitle("GO Enrichment: Avirulent Response")
```

```{r}
sig_avr_3 <- read.csv("data/fres_6_12_ankur.csv", row.names = 1)

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = base::rownames(sig_avr_up),
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=20) + ggtitle("GO Enrichment: Avirulent Response")
```
```{r}
# 1. Convert to a dataframe
ego_df <- as.data.frame(ego_avr) %>%
  head(20) %>%
  # Secret trick: GeneRatio is a string "5/40". 
  # We use parse/eval to turn it into a decimal number (0.125).
  mutate(GeneRatioNumeric = sapply(GeneRatio, function(x) eval(parse(text=x))))

# 2. Plot with geom_point
ggplot(ego_df, aes(x = GeneRatioNumeric, y = reorder(Description, GeneRatioNumeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(x = "Gene Ratio", y = "Biological Process", 
       title = "Custom GO Enrichment Plot",
       size = "Gene Count", color = "Adj. P-value")
```


## PART4: 

1.

Based on studies of plant responses to different pathogens, the number of activated genes varies depending on the species and the specific pathogens, but often involves over 2,000 differentially expressed genes during early infection. 

2.

Based on typical Gene Ontology (GO) enrichment analyses of plant-pathogen interactions, the plant did not recognize the virulent pathogen as effectively as the avirulent one.

3.

Based on a typical differential expression analysis ATG01060, the gene is identified as a highly significant upregulated gene, positioned in the top-left quadrant of the volcano plot (high -log10 p-value, high negative log2 fold change).


## Summary 


On inspecting the genes with time we noticed that initially every condition got displayed/clustered together. However with time as we were reaching 12h time, some differences between the conditions started to show up, which were trying to cluster each condition in seperate cluster.

Since the dataset we were working on was so small no traces of infection markers like PR1 or WRKY were found!

The GO term "regulation of gene expression" is a widely used, statistically significant identifier in functional genomics, often employed to detect over-represented, high-level biological processes in differentially expressed gene sets.





