---
title: "transcriptomics_day3_ankur"
format: html
---
```{r}
#| eval: False
  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')

  BiocManager::install('EnhancedVolcano')
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}
# Define the required packages
pkgs <- c("tidyverse", "DESeq2", "DT", "pheatmap", "RColorBrewer", "EnhancedVolcano", "ape")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)
```
# Merging

```{r}
# 1. Load all files available in your group
my_counts      <- read.csv("data/mock_vs_avr.csv", row.names = 1)
partner_1_counts <- read.csv("data/MockVsVirulent_Specific_Counts.csv", row.names = 1)
partner_2_counts <- read.csv("data/Virulent_Avirulent_Counts.csv", row.names = 1) # Optional for Trios

# 2. Start with your counts as the base Master Matrix
# We convert row names to a column named "GeneID" to enable the join
counts_master <- my_counts %>% rownames_to_column("GeneID")

# 3. Merge Partner A - The "Safe-Merge" Pipe
# any_of() checks the column names of our master table and 
# removes them from the partner's table BEFORE joining.
counts_master <- partner_1_counts %>% 
  rownames_to_column("GeneID") %>%
  select(-any_of(colnames(counts_master)), GeneID) %>%
  inner_join(counts_master, by = "GeneID")

# 4. Merge Partner B (Only if you are a group of 3)
counts_master <- partner_2_counts %>% 
  rownames_to_column("GeneID") %>%
  select(-any_of(colnames(counts_master)), GeneID) %>%
  inner_join(counts_master, by = "GeneID")

# 5. Finalize the matrix
counts_master <- counts_master %>% column_to_rownames("GeneID")

# 6. Sanity Check: 
# Regardless of group size, you should see exactly 18 columns 
# (6 Mock, 6 Virulent, 6 Avirulent)
ncol(counts_master)


```
```{r}
write.csv(counts_master, "data/mock_vs_avr_vs_vir.csv", row.names = TRUE)
```


## Creating the colData

```{r}
# Create a metadata frame
# Ensure the levels are set so 'Mock' is the reference (the baseline)
colData <- data.frame(
  row.names = colnames(counts_master),
  Condition = factor(rep(c("Virulent", "Avirulent", "Mock"), each = 6), 
                     levels = c("Mock", "Virulent", "Avirulent")),
  Time = factor(rep(rep(c("12h", "1h", "6h"), each = 2), 3))
)
```

```{r}
colData$Condition
```

## Task: Metadata Expansion

```{r}
colData$type <- factor(rep("single-read", nrow(colData)))

print(colData)

all(rownames(colData) == colnames(counts_master))
```
## DESeq2

```{r}
# Create the dds object
dds <- DESeqDataSetFromMatrix(countData = round(counts_master),
                              colData = colData,
                              design = ~ Condition)

```


## Building the Dendrogram

```{r}
# 1. Transform the data to stabilize variance
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
vsd_mat <- assay(vsd) # Extract the matrix of transformed values

# 2. Calculate the distance between samples
# We transpose the matrix with t() because dist() calculates distance between rows
sample_dist <- dist(t(vsd_mat))

# 3. Perform Hierarchical Clustering
clusters <- hclust(sample_dist)

# 4. Plot the results using the ape package for a cleaner phylogenetic look
# We convert the hclust object to a phylo object for professional plotting.
# We map 'Condition' to 'tip.label' so the tree is easy to interpret.
plot.phylo(as.phylo(clusters), 
           tip.label = colData$Condition, 
           type = "p", 
           edge.col = "grey", 
           edge.width = 2,
           no.margin = TRUE)
```
If the data is big instead of using varianceStabilizingTransformation us vst. In that case there is possibility of more clear clustering.

1. According to the diagram above, No 2 same condition Mock able to get cluster together this maybe due to small small dataset or its because of small dataset.

2. No, there is no clear difference according to condition, however difference based on time is slightly noticeable.

3. In the experiment Virulent sample sitting in the middle of the Mock branch, suggest they are similar to each other. 


## The Main Event - Differential Expression


```{r}
# Assign the output to a new object to keep your original dds safe
dds_results <- DESeq(dds)
```

```{r}
# Extract results for Avirulent vs. Mock
res_avr_mock <- results(dds_results, contrast = c("Condition", "Avirulent", "Mock"), alpha = 0.2)

# Print the first few rows to see the table
print(res_avr_mock)
```

```{r}
# See a high-level summary of the results
summary(res_avr_mock)
```

```{r}
res_vir_mock <- results(dds_results, contrast = c("Condition", "Virulent", "Mock"), alpha = 0.2)

# Print the first few rows to see the table
print(res_vir_mock)

```

```{r}
summary(res_vir_mock)
```

```{r}
res_vir_avr <- results(dds_results, contrast = c("Condition", "Virulent", "Avirulent"), alpha = 0.2)

# Print the first few rows to see the table
print(res_vir_avr)

```

```{r}
summary(res_vir_avr)
```

```{r}
# Create the dds2 object
dds2 <- DESeqDataSetFromMatrix(countData = round(counts_master),
                              colData = colData,
                              design = ~ Time)
```

```{r}
# 1. Transform the data to stabilize variance
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
vsd_mat <- assay(vsd) # Extract the matrix of transformed values

# 2. Calculate the distance between samples
# We transpose the matrix with t() because dist() calculates distance between rows
sample_dist <- dist(t(vsd_mat))

# 3. Perform Hierarchical Clustering
clusters <- hclust(sample_dist)

# 4. Plot the results using the ape package for a cleaner phylogenetic look
# We convert the hclust object to a phylo object for professional plotting.
# We map 'Condition' to 'tip.label' so the tree is easy to interpret.
plot.phylo(as.phylo(clusters), 
           tip.label = colData$Condition, 
           type = "p", 
           edge.col = "grey", 
           edge.width = 2,
           no.margin = TRUE)
```
```{r}
# Assign the output to a new object to keep your original dds safe
dds_results <- DESeq(dds2)
```

```{r}
# Extract results for
res_1_12 <- results(dds_results, contrast = c("Time", "1h", "12h"), alpha = 0.2)

# Print the first few rows to see the table
print(res_1_12)
```
```{r}
# Extract results for
res_6_12 <- results(dds_results, contrast = c("Time", "6h", "12h"), alpha = 0.2)

# Print the first few rows to see the table
print(res_6_12)
```
```{r}
# Extract results for
res_1_6 <- results(dds_results, contrast = c("Time", "1h", "6h"), alpha = 0.2)

# Print the first few rows to see the table
print(res_1_6)
```


```{r}
summary(res_1_12)
```
```{r}
summary(res_1_6)
```

```{r}
summary(res_6_12)
```
## Filtering and Ranking

```{r}
# Function to filter and count DE genes
filter_and_count <- function(res_obj, comparison_name, fc_threshold = 2) {
  # 1. Remove rows with NAs in padj
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  
  # 2. Apply filters: |log2FC| >= log2(fc_threshold) and padj < 0.2
  # Note: 0.2 matches the alpha we set in our results() calls
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold) & res_filtered$padj < 0.2, ]
  
  # 3. Count up and down regulated
  up_regulated <- sum(sig_genes$log2FoldChange > 0)
  down_regulated <- sum(sig_genes$log2FoldChange < 0)
  
  # Return as a clean data frame
  return(data.frame(
    Comparison = comparison_name,
    Up_regulated = up_regulated,
    Down_regulated = down_regulated
  ))
}
```

```{r}
results_summary <- rbind(
  filter_and_count(res_vir_mock, "Vir vs Mock"),
  filter_and_count(res_avr_mock, "Avr vs Mock"),
  filter_and_count(res_vir_avr, "Vir vs Avr")
)

# Inspect the final summary table
print(results_summary)
```

```{r}
write.csv(results_summary, "data/results_summary_ankur.csv", row.names = TRUE)
```


```{r}
results_summary2 <- rbind(
  filter_and_count(res_1_6, "1h vs 6h"),
  filter_and_count(res_1_12, "1h vs 12h"),
  filter_and_count(res_6_12, "6h vs 12h")
)

# Inspect the final summary table
print(results_summary2)
```

```{r}
results_summary |>
pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count")
```

```{r}
results_summary2 |>
pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count")
```
```{r}
write.csv(results_summary2, "data/results_summary2_ankur.csv", row.names = TRUE)
```

## Graph

```{r}

plot_data <- results_summary %>%
  pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count") %>%
  mutate(Regulation = factor(Regulation, levels = c("Up_regulated", "Down_regulated")))

ggplot(plot_data, aes(x = Comparison, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "dodge")
```

```{r}
plot_data <- results_summary2 %>%
  pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count") %>%
  mutate(Regulation = factor(Regulation, levels = c("Up_regulated", "Down_regulated")))

ggplot(plot_data, aes(x = Comparison, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "dodge")
```

Reflction 

1. How does merging all 18 samples help DESeq2 estimate “dispersion” more robustly?
Ans: It will help eleminating the posibility of batch standardising/normalising

2. Why is a 2-fold change (log2FoldChange of 1) often the minimum threshold for biological interest?
Ans: A 2-fold change (log2FoldChange of 1 or -1) is commonly used as a minimum threshold in differential expression analysis to balance statistical significance with biological relevance and experimental noise reduction. While lower thresholds can be used, a 2-fold change is traditionally regarded as a reliable indicator of a robust change in gene expression that is likely to impact cellular phenotypes

