---
title: "transcriptomics_day3_kimia_p1"
author: "Kimia"
format: html
editor: visual
toc: true
---

# **The basics of transcriptomics pt3**

## **Part 0: Computational Environment**

```{r}
BiocManager::install("EnhancedVolcano")
# Define the required packages
pkgs <- c("tidyverse", "DESeq2", "DT", "pheatmap", "RColorBrewer", "EnhancedVolcano", "ape")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)
```

## **Part 1: The Great Merger**

```{r}
count_master <- read.csv("data/mock_vs_avr_vs_vir.csv", row.names = 1)
```

## **Part 2: The Experimental Design Matrix**

### **1. Creating the colData**

```{r}
# Create a metadata frame
# Ensure the levels are set so 'Mock' is the reference (the baseline)
colData <- data.frame(
  row.names = colnames(count_master),
  Condition = factor(rep(c("Virulent", "Avirulent", "Mock"), each = 6), 
                     levels = c("Virulent", "Avirulent", "Mock")),
  Time = factor(rep(rep(c("12h", "1h", "6h"), each = 2), 3))
)
```

### **2. Task: Metadata Expansion**

```{r}
colData <- colData %>%
  mutate(type = factor(rep("single-read")))

print(colData)
```

```{r}
all(rownames(colData) == colnames(count_master))
```

## **Part 3: Introduction to DESeq2**

### **2. Constructing the DESeq Object**

```{r}
# Create the dds object
dds <- DESeqDataSetFromMatrix(countData = round(count_master),
                              colData = colData,
                              design = ~ Condition)

dds2 <- DESeqDataSetFromMatrix(countData = round(count_master),
                              colData = colData,
                              design = ~ Time)
```

## **Part 4: Exploratory Data Analysis: Hierarchical Clustering**

### **3. Task: Building the Dendrogram**

```{r}
# 1. Transform the data to stabilize variance
vsd <- varianceStabilizingTransformation(dds, blind =FALSE)
vsd_mat <- assay(vsd) # Extract the matrix of transformed values

# 2. Calculate the distance between samples
# We transpose the matrix with t() because dist() calculates distance between rows
sample_dist <- dist(t(vsd_mat))

# 3. Perform Hierarchical Clustering
clusters <- hclust(sample_dist)

# 4. Plot the results using the ape package for a cleaner phylogenetic look
# We convert the hclust object to a phylo object for professional plotting.
# We map 'Condition' to 'tip.label' so the tree is easy to interpret.
plot.phylo(as.phylo(clusters), 
           tip.label = colData$Condition, 
           type = "p", 
           edge.col = "grey", 
           edge.width = 2,
           no.margin = TRUE)
```

-   Look at the dendrogram. Do the replicates for the same condition (e.g., the two `Mock` samples at 1h) cluster together on the same small branches?

    A lot of the clusters early in the infection are clustered together based on time and not condition, which the later hours' clusters are based on condition. Early and late groups.

-   Do you see a clear split between the `Mock` group and the infected groups?

    We can see some of the mock groups being clustered together with the infected groups, e.g. M6B and A6B since the initial injury looks the same.

-   If a sample is “out of place” (e.g., a Virulent sample sitting in the middle of the Mock branch), what might that suggest about the experiment?

    That maybe the infection takes some time to take place and it takes time for the plant to express its defense genes. There are differences in the data.

```{r}
vsd <- varianceStabilizingTransformation(dds2, blind =FALSE)
vsd_mat <- assay(vsd) # Extract the matrix of transformed values

# 2. Calculate the distance between samples
# We transpose the matrix with t() because dist() calculates distance between rows
sample_dist <- dist(t(vsd_mat))

# 3. Perform Hierarchical Clustering
clusters <- hclust(sample_dist)

# 4. Plot the results using the ape package for a cleaner phylogenetic look
# We convert the hclust object to a phylo object for professional plotting.
# We map 'Condition' to 'tip.label' so the tree is easy to interpret.
plot.phylo(as.phylo(clusters), 
           tip.label = colData$Time, 
           type = "p", 
           edge.col = "grey", 
           edge.width = 2,
           no.margin = TRUE)
```

## **Part 5: The Main Event - Differential Expression**

### **1. Running the Pipeline - our null hypothesis is that all genes are expressed the same.** 

```{r}
# Assign the output to a new object to keep your original dds safe
dds_results <- DESeq(dds)

dds_results2 <- DESeq(dds2)
```

### **2. Extracting Specific Results**

alpha is usually the p-value. Here alpha represents the your target **False Discovery Rate (FDR)**. The proportion of false positives among all significant results. 20% of our significant genes with (0.05% p value for being differentially expressed) might be false positives. Used the p-value to get significant ones and estimated the proportions of falsely discovered genes.

```{r}
# Extract results for Avirulent vs. Mock
res_avr_mock <- results(dds_results, contrast = c("Condition", "Avirulent", "Mock"), alpha = 0.2)

res_1h_12h <- results(dds_results2, contrast = c("Time", "1h", "12h"), alpha = 0.2)

# Print the first few rows to see the table
print(res_avr_mock)
print(res_1h_12h)
```

```{r}
# See a high-level summary of the results
summary(res_1h_12h)
```

```{r}
res_vir_mock <- results(dds_results, contrast = c("Condition", "Virulent", "Mock"), alpha = 0.2)

res_vir_avr <- results(dds_results, contrast = c ("Condition", "Virulent", "Avirulent"), alpha = 0.2)

res_1h_6h <-results(dds_results2, contrast = c("Time", "1h", "6h"), alpha = 0.2)

res_6h_12h <- results(dds_results2, contrast = c("Time", "6h", "12h"), alpha = 0.2)

summary(res_vir_mock)
summary(res_vir_avr)
summary(res_1h_6h)
summary(res_6h_12h)
```

## **Part 6: Filtering and Ranking**

### **2. Automation with Custom Functions**

```{r}
# Function to filter and count DE genes
filter_and_count <- function(res_obj, comparison_name, fc_threshold = 2) {
  # 1. Remove rows with NAs in padj
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  
  # 2. Apply filters: |log2FC| >= log2(fc_threshold) and padj < 0.2
  # Note: 0.2 matches the alpha we set in our results() calls
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold) & res_filtered$padj < 0.2, ]
  
  # 3. Count up and down regulated
  up_regulated <- sum(sig_genes$log2FoldChange > 0)
  down_regulated <- sum(sig_genes$log2FoldChange < 0)
  
  # Return as a clean data frame
  return(data.frame(
    Comparison = comparison_name,
    Up_regulated = up_regulated,
    Down_regulated = down_regulated
  ))
}

```

### **3. Generating the Global Summary**

```{r}
results_summary <- rbind(
  filter_and_count(res_1h_12h, "1h vs 12h"),
  filter_and_count(res_1h_6h, "1h vs 6h"),
  filter_and_count(res_6h_12h, "6h vs 12h")
)

results_summary2 <- rbind(
  filter_and_count(res_vir_mock, "Vir vs Mock"),
  filter_and_count(res_avr_mock, "Avr vs Mock"),
  filter_and_count(res_vir_avr, "Vir vs Avr")
)

# Inspect the final summary table
print(results_summary2)

# Inspect the final summary table
print(results_summary)
```

```{r}
# Step 1: Pivot the data to long format
plot_data <- results_summary %>%
  pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count") %>%
  mutate(Regulation = factor(Regulation, levels = c("Up_regulated", "Down_regulated")))

# Step 2: Create a Bar Plot
# Hint: Use aes(x = Comparison, y = Count, fill = Regulation) 
# and geom_bar(stat = "identity", position = "dodge")

ggplot (plot_data, aes (x = Comparison, y = Count, fill = Regulation)) +
         geom_bar(stat = "identity", position = "dodge")
```

# **Final Submission**
